Equalizer Programming Guide

### 2008-08-25 Developer/ProgrammingGuide/paper.tex
Programming Guide: Update for OSWindow changes.
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\date{
  \vspace{2cm}
  \begin{figure}[ht]
    \hspace{-4.5cm}\includegraphics[width=22cm]{images/teaser.pdf}
  \end{figure}
  \vspace{1cm}
%  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
  Version 1.3.7 for Equalizer 0.5.4+svn\\[\medskipamount]
  \today
}

@@

\clearpage
\thispagestyle{empty}
\tableofcontents
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Version for Equalizer 0.5\\
    1.3.1   & May 02, 2008 & Add \sref{sStatistics}, improved images based on 1.2 print\\
    1.3.2   & May 05, 2008 & Add network distribution on eqPly model data\\
    1.3.3   & May 10-19, 2008 & add \aref{aFileFormat}\\
    1.3.4   & July 3, 2008 & Update \sref{sStatistics}\\
    1.3.5   & July 7, 2008 & Update after namespace naming cleanup\\
    1.3.6   & July 20, 2008 & Add accum and samples to \aref{aFileFormat}, add \sref{sDrawableConfig}\\
    1.3.7   & Aug 22, 2008 & Update \sref{sEqplyWIndow} with OSWindow changes (event handling updates still missing)\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

@@

The default implementation provides a convenient, easily accessible
event framework, while allowing all necessary customizations. It gathers
all events in the main thread of the application, so that the developer only
has to implement \textsf{Config::processEvent} to update its data based
on the pre-processed, generic keyboard and mouse events. It is very easy
to use and similar to an GLUT-based implementation.

@@

During window and pipe initialization the event handling is set up. For
both entities, \textsf{initEvent\-Handler} is called to register the
pipe or window with an event handler. This method may be overwritten to
use a custom event handler, or to not install an event handler at all in
order to disable event handling. Likewise, \textsf{exitEventHandler} is
called to de-initialize event handling.

An event handler consists of two parts: the generic base class providing
the interface and generic functions, and the window-system-specific
part providing the actual implementation.

Event handling is initialized whenever a new window-system-specific
pipe or window handle is set. First, \textsf{exitEventHandler} is called
to de-initialize event handling for the old handle (if set), and then
\textsf{initEvent\-Handler} is called for the new handle.

AGL uses an event handler singleton, GLX uses one event handler per pipe
and WGL uses one event handler per window.

@@

In order to dispatch the events, Equalizer 'pumps' the native events. On
WGL and GLX, this has happens on each thread with windows, whereas on
AGL it has to happen only on the main thread. By default, Equalizer
pumps these events automatically for the application in-between
executing task methods.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
  {\caption{\small\label{fEventProcessing}Event Processing}}
  \vspace{-3ex}
\end{wrapfigure}
Events are received by an event handler. The event handler finds the
\textsf{eq::Window} associated to the event. It then creates a generic
\textsf{WindowEvent}, which holds important event data in an independent
format. The original event is attached to the generic window event.

The event handler then passes the window event to
\textsf{Window::processEvent}, which is responsible for either handling
the event locally, or for translating it into a generic
\textsf{ConfigEvent}. The config events are send to the application
thread using \textsf{Config::sendEvent}.

If the event was processed by \textsf{processEvent}, the function has to
return \textsf{true}. If \textsf{false} is returned, the event will be
passed to a previously installed, window-system-specific event handling
function. The default implementation of \textsf{Window::pro\-cess\-Event}
passes most events on to the application.

Events send using \textsf{Config::sendEvent} are queued in the
application thread. After a frame has been finished,
\textsf{Config::finishFrame} calls \textsf{Config::handleEvents}. The
default implementation of this method provides non-blocking event
processing, that is, it calls \textsf{Config::handleEvent} for each
queued event. By overriding \textsf{handleEvents}, event-driven
execution can be implemented.

Later Equalizer versions will introduce \textsf{Pipe::pro\-cess\-Event} and
\textsf{PipeEvent} to communicate pipe-specific events, e.g., monitor
resolution changes.

@@

Each entity uses \textsf{waitFrame\-Started} to block on the parent's
\textsf{startFrame}, e.g., \textsf{Pipe::\-frame\-Start} calls
\textsf{Node::wait\-Frame\-Started} to wait for the corresponding
\textsf{Node::start\-Fra\-me}. This explicit synchronization allows to
update non-critical data before synchronizing with
\textsf{waitFrameStarted}, or after unlocking using
\textsf{start\-Fra\-me}. \fig{fFrameSync} illustrates this
synchronization model.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
  {\caption{\small\label{fFrameSync}Per-Node Frame Synchronization}}
\end{wrapfigure}
\fig{fFrameSync} illustrates the local frame synchronization. By
removing the calls to \textsf{waitFrameLocal}, the node thread can
disable the synchronization with the pipe threads, which causes the node
operations to overlap with the rendering.

@@

In direct send compositing, each rendering channel is also responsible
for the sort-last composition of one screen-space tile. He receives
the framebuffer pixels for his tile from all the other channels. The
size of one tile decreases linearly with the number of source channels,
which keeps the total amount of pixel data per channel constant.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
  {\caption{\small\label{fDirectSend}Parallel Direct Send Compositing}}
\end{wrapfigure}
\fig{fDirectSend} illustrates this algorithm for three channels. The
Equalizer website contains a
presentation\footnote{\link{http://www.equalizergraphics.com/documents/EGPGV07.pdf}}
explaining and comparing this algorithm with the binary-swap algorithm.

@@

Readback and assemble operations on frames and images are designed to be
asynchronous. They have a start and finish method for both readback and
assemble to allow the initiation and synchronization of the operation.
Currently, only synchronous readback and assembly using
\textsf{glReadPixels} and \textsf{glDrawPixels} is implemented in the
respective start method of the image. Later versions of Equalizer will
implement asynchronous pixel transfers.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/assembly.pdf}
  {\caption{\small\label{fAssembly}Hierarchy of assembly classes}}
\end{wrapfigure}
The offset of input and output frames characterizes the position of the
frame data with respect to the framebuffer, that is, the
\textbf{window's} lower-left corner. For output frames this is the
position of the channel with respect to the window.

@@

For recomposition, the 2D frame buffer contents are blended together to
form a seamless picture. For correct blending, the frames are ordered in
the same back-to-front order as the slices used for rendering, and use the
same blending parameters. Simplified, the frame buffer images are
`thick' slices which are `rendered' by writing their content to the
destination frame buffer using the correct order.

>>>>

\date{
  \vspace{2cm}
  \begin{figure}[ht]
    \hspace{-4.5cm}\includegraphics[width=22cm]{images/teaser.pdf}
  \end{figure}
  \vspace{1cm}
%  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
  Version 1.3.8 for Equalizer 0.5.4+svn\\[\medskipamount]
  \today
}

@@

\clearpage
\thispagestyle{empty}
\tableofcontents
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Version for Equalizer 0.5\\
    1.3.1   & May 02, 2008 & Add \sref{sStatistics}, improved images based on 1.2 print\\
    1.3.2   & May 05, 2008 & Add network distribution on eqPly model data\\
    1.3.3   & May 10-19, 2008 & add \aref{aFileFormat}\\
    1.3.4   & July 3, 2008 & Update \sref{sStatistics}\\
    1.3.5   & July 7, 2008 & Update after namespace naming cleanup\\
    1.3.6   & July 20, 2008 & Add accum and samples to \aref{aFileFormat}, add \sref{sDrawableConfig}\\
    1.3.7   & Aug 22, 2008 & Update \sref{sEqplyWIndow} with OSWindow changes\\
    1.3.8   & Aug 25, 2008 & Update \sref{sEventHandling} with OSWindow changes\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

@@

The default implementation provides a convenient, easily accessible
event framework, while allowing all necessary customizations. It gathers
all events from all node processes in the main thread of the
application, so that the developer only has to implement
\textsf{Config::processEvent} to update its data based on the
pre-processed, generic keyboard and mouse events. It is very easy to use
and similar to an GLUT-based implementation.

@@

@@

In order to dispatch events, Equalizer 'pumps' the native events. On
WGL and GLX, this happens on each thread with windows, whereas on
AGL it happens only on the main thread. By default, Equalizer
pumps these events automatically for the application in-between
executing task methods.

@@

Events are received by an event handler. The event handler finds the
\textsf{eq::OSWindow} associated to the event. It then creates a generic
\textsf{WindowEvent}, which holds the event data in an independent
format. The original native event and this generic \textsf{WindowEvent}
form the \textsf{OSWindowEvent}, which is passed to the concrete
\textsf{OSWindow} for processing.

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
  {\caption{\small\label{fEventProcessing}Event Processing}}
\end{wrapfigure}
The purpose of the \textsf{OSWindow} processing method,
\textsf{processEvent}, is to perform window-system specific event
processing. For example, \textsf{AGLWindow::processEvent} calls
\textsf{aglUpdateContext} whenever a resize event is received. For
further processing, the \textsf{WindowEvent} is passed on to
\textsf{Window::processEvent}. This \textsf{WindowEvent} no longer contains
the native event.

\textsf{Window::processEvent} is responsible for either handling
the event locally, or for translating it into a generic
\textsf{ConfigEvent}. The config events are send to the application
thread using \textsf{Config::sendEvent}.

Events send using \textsf{Config::sendEvent} are queued in the
application thread. After a frame has been finished,
\textsf{Config::finishFrame} calls \textsf{Config::handleEvents}. The
default implementation of this method provides non-blocking event
processing, that is, it calls \textsf{Config::handleEvent} for each
queued event. By overriding \textsf{handleEvents}, event-driven
execution can easily be implemented.

If the event was processed, \textsf{processEvent} has to return
\textsf{true}. If \textsf{false} is returned to the event handler, the
event will be passed to the previously installed, window-system-specific
event handling function.

\fig{fEventProcessing} shows the overall data flow of an event.

\subsubsection{Default Implementation}

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/eventUML.pdf}
  {\caption{\small\label{fEventUML}UML Diagram for Event Handling Classes}}
\end{wrapfigure}
Currently, Equalizer provides an \textsf{AGLEvent\-Handler},
\textsf{GLXEventHandler} and \textsf{WGL\-Event\-Handler}, which handle
events for an \textsf{AGLWindowIF}, \textsf{GLXWindowIF} and
\textsf{WGLWindowIF}, respectively. \fig{fEventUML} illustrates the
class hierarchy for event processing.

The \textsf{OSWindow} implementation is responsible for setting up and
de-initializing event handling.

Carbon events issued for AGL windows are processed from the process main
thread. One \textsf{AGLEventHandler} is used. When a Carbon window is
set on an \textsf{AGLWindow}, the window is registered with the
\textsf{AGLEventHandler}. The event handler installs a Carbon event
handler for all important events. The event handler uses an
\textsf{AGLWindowEvent} to pass the Carbon \textsf{EventRef} to
\textsf{AGLWindowIF::processEvent}. During window exit, the installed
Carbon handler is removed when the window is deregistered with the
\textsf{AGLEventHandler}. No event handling is set up for PBuffers.

GLX uses one X11 Display connection for all windows on a GPU, which is
managed by the \textsf{eq::Pipe}. For each thread, one
\textsf{GLXEventHandler} is allocated. The pipe registers the display
connection with this per-thread event handler. Events for windows
created using this display connection are automatically received by the
event handler. No per-window event handling setup needs to be
done. PBuffers are handled in the same way as window drawables.

The glX event handler iterates over all windows of the registered pipe
to find the originating \textsf{GLXWindowIF}, which is then used to
process the event in a fashion similar to AGL. The
\textsf{GLXWindowEvent} passes the \textsf{XEvent} to
\textsf{GLXWindowIF::processEvent}.

Each \textsf{WGLWindow} allocates one \textsf{WGLEventHandler} when the
window handle is set. The WGLEventHandler passes the native event
parameters \textsf{uMsg}, \textsf{wParam} and \textsf{lParam} to
\textsf{WGLWindowIF::processEvent} as part of the
\textsf{WGLWindowEvent}. No event handling is set up for PBuffers.

Later Equalizer versions will introduce \textsf{Pipe::pro\-cess\-Event} and
\textsf{PipeEvent} to communicate pipe-specific events, e.g., monitor
resolution changes. It is also likely that an \textsf{OSPipe}
abstraction, similar to the \textsf{OSWindow}, will be created to
separate the window system specific code from the \textsf{eq::Pipe}.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
  {\caption{\small\label{fFrameSync}Per-Node Frame Synchronization}}
\end{wrapfigure}
Each entity uses \textsf{waitFrame\-Started} to block on the parent's
\textsf{startFrame}, e.g., \textsf{Pipe::\-frame\-Start} calls
\textsf{Node::wait\-Frame\-Started} to wait for the corresponding
\textsf{Node::start\-Fra\-me}. This explicit synchronization allows to
update non-critical data before synchronizing with
\textsf{waitFrameStarted}, or after unlocking using
\textsf{start\-Fra\-me}. \fig{fFrameSync} illustrates this
synchronization model.

@@

\fig{fFrameSync} illustrates the local frame synchronization. By
removing the calls to \textsf{waitFrameLocal}, the node thread can
disable the synchronization with the pipe threads, which causes the node
operations to overlap with the rendering.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
  {\caption{\small\label{fDirectSend}Parallel Direct Send Compositing}}
\end{wrapfigure}
In direct send compositing, each rendering channel is also responsible
for the sort-last composition of one screen-space tile. He receives
the framebuffer pixels for his tile from all the other channels. The
size of one tile decreases linearly with the number of source channels,
which keeps the total amount of pixel data per channel constant.

@@

\fig{fDirectSend} illustrates this algorithm for three channels. The
Equalizer website contains a
presentation\footnote{\link{http://www.equalizergraphics.com/documents/EGPGV07.pdf}}
explaining and comparing this algorithm with the binary-swap algorithm.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/assembly.pdf}
  {\caption{\small\label{fAssembly}Hierarchy of assembly classes}}
\end{wrapfigure}
Readback and assemble operations on frames and images are designed to be
asynchronous. They have a start and finish method for both readback and
assemble to allow the initiation and synchronization of the operation.
Currently, only synchronous readback and assembly using
\textsf{glReadPixels} and \textsf{glDrawPixels} is implemented in the
respective start method of the image. Later versions of Equalizer will
implement asynchronous pixel transfers.

The offset of input and output frames characterizes the position of the
frame data with respect to the framebuffer, that is, the
\textbf{window's} lower-left corner. For output frames this is the
position of the channel with respect to the window.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/volResult.png}
  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
  \vspace{-1em}
\end{wrapfigure}
For recomposition, the 2D frame buffer contents are blended together to
form a seamless picture. For correct blending, the frames are ordered in
the same back-to-front order as the slices used for rendering, and use the
same blending parameters. Simplified, the frame buffer images are
`thick' slices which are `rendered' by writing their content to the
destination frame buffer using the correct order.
